---
title: "How Reliable Is the Delta Method CI? Bootstrap Validation and Sample Size Planning"
draft: true
draft-mode: unlinked
date: "2026-07-15"
categories: ["brand-lift", "statistics"]
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

The [previous post](/posts/bls-delta-method-rr/) derived a log-transform delta method CI for the risk ratio $RR = \frac{p_1}{p_0}$:

$$\left( \widehat{RR} \cdot e^{-z_{1-\alpha/2} \cdot SE}, \quad \widehat{RR} \cdot e^{+z_{1-\alpha/2} \cdot SE} \right)$$

where $SE = \sqrt{\frac{1}{x_1} - \frac{1}{n_1} + \frac{1}{x_0} - \frac{1}{n_0}}$.

It is closed-form, computable in SQL, and always positive. But it is based on a first-order Taylor approximation. Brand Lift Study (BLS) campaigns run at finite sample sizes, and the baseline rate $p_0$ can be low. How does the interval actually perform?

This post answers that question by simulation. A bootstrap percentile CI serves as the comparison method: it does not depend on the normal approximation that the delta method requires. Both are evaluated by coverage, the proportion of simulated datasets where the CI contains the true $RR$.

The short version: when the expected number of positive control responses $\mathbb{E}[x_0] = n \cdot p_0$ is 20 or more, the delta method CI achieves its nominal 95% coverage. Below that, it becomes conservative, producing intervals wider than the stated level implies. The second half of the post derives the minimum per-group sample size $n_{\min}$ required to detect a given lift with a target power.

Throughout, $n_1 = n_0 = n$, for simplicity.

## Bootstrap percentile CI for RR

The bootstrap percentile CI does not require the normal approximation underlying the delta method. It is a parametric bootstrap: resample from the estimated binomial distributions and take quantiles.

Given observed counts $(x_1, n, x_0, n)$:

1.  Compute $\hat{p}_1 = \frac{x_1}{n}$ and $\hat{p}_0 = \frac{x_0}{n}$.
2.  For $b = 1, \dots, B$: draw $x_1^* \sim \text{Binomial}(n, \hat{p}_1)$ and $x_0^* \sim \text{Binomial}(n, \hat{p}_0)$ independently. If $x_0^* = 0$, discard the replicate ($\widehat{RR}^*$ is undefined). Otherwise compute $\widehat{RR}^*_b = (x_1^*/n) / (x_0^*/n)$.
3.  The $(1-\alpha)$ CI is $\left( \widehat{RR}^*_{(\alpha/2)}, \; \widehat{RR}^*_{(1-\alpha/2)} \right)$.

## Coverage simulation

### Why fix RR?

The simulation varies $p_0$ and $n$ but holds $RR$ fixed. Here is why.

$\log(\widehat{RR}) = \log(\hat{p}_1) - \log(\hat{p}_0)$, and the two terms are independent. The normal approximation to their difference is only as good as the worse of the two individual approximations.

The delta method approximation to $\log(\hat{p})$ depends on the expected count $np = \mathbb{E}[x]$. Large $np$ means $\hat{p}$ is approximately normal and $\log(\hat{p})$ inherits that normality. Small $np$ means $\hat{p}$ is discrete and right-skewed, and the approximation breaks down.

With $n_1 = n_0 = n$ and $RR > 1$, $p_1 = RR \cdot p_0 > p_0$, so $np_1 > np_0$ always. The exposed group always has more expected positive responses than control. The approximation to $\log(\hat{p}_0)$ is always the weaker link.

The accuracy of $\log(\hat{p}_0)$'s approximation depends on $np_0$ alone. $RR$ affects $p_1$ but not $p_0$. Changing $RR$ does not shift the threshold where the delta method starts to deviate.

The main simulation uses $RR = 1.20$, a moderate lift typical of BLS campaigns (I think). One confirmation condition at $RR = 1.50$ checks the $RR$-independence numerically.

### Design

The key quantity is $\mathbb{E}[x_0] = n \cdot p_0$. The grid varies both $p_0$ and $n$ to generate a range of $\mathbb{E}[x_0]$ values.

| Parameter            | Values                                  |
|----------------------|-----------------------------------------|
| $RR$                 | $1.20$ (primary), $1.50$ (confirmation) |
| $p_0$                | $0.02, \; 0.05, \; 0.10, \; 0.20$       |
| $n \; (= n_1 = n_0)$ | $100, \; 500, \; 2{,}000$               |
| Replications         | $K = 10{,}000$                          |
| Confidence level     | $95\%$                                  |

Four $p_0$ values span BLS baseline rates at half-log spacing. Three $n$ values produce 12 conditions with 10 unique $\mathbb{E}[x_0]$ from 2 to 400:

|              | $n = 100$ | $n = 500$ | $n = 2{,}000$ |
|--------------|-----------|-----------|---------------|
| $p_0 = 0.02$ | 2         | 10        | 40            |
| $p_0 = 0.05$ | 5         | 25        | 100           |
| $p_0 = 0.10$ | 10        | 50        | 200           |
| $p_0 = 0.20$ | 20        | 100       | 400           |

Two pairs share the same $\mathbb{E}[x_0]$: $\mathbb{E}[x_0] = 10$ from $(0.02,\, 500)$ and $(0.10,\, 100)$; $\mathbb{E}[x_0] = 100$ from $(0.05,\, 2{,}000)$ and $(0.20,\, 500)$. These test whether $\mathbb{E}[x_0]$ alone determines coverage.

Each replication draws $x_1 \sim \text{Binomial}(n, p_1)$ and $x_0 \sim \text{Binomial}(n, p_0)$ from the true parameters, computes both CIs, and records whether each contains the true $RR$. Replications where either count is zero are skipped.

### Results

```{r}
#| label: sim-functions
#| include: false

bls_rr_ci <- function(x1, n1, x0, n0, alpha = 0.05) {
  p1 <- x1 / n1
  p0 <- x0 / n0
  rr <- p1 / p0
  se_log <- sqrt(1 / x1 - 1 / n1 + 1 / x0 - 1 / n0)
  z <- qnorm(1 - alpha / 2)
  exp(log(rr) + c(-1, 1) * z * se_log)
}

bls_bootstrap_ci <- function(x1, n1, x0, n0, B = 5000, alpha = 0.05) {
  p1_hat <- x1 / n1
  p0_hat <- x0 / n0
  x1_star <- rbinom(B, n1, p1_hat)
  x0_star <- rbinom(B, n0, p0_hat)
  
  valid <- x0_star > 0 & x1_star > 0
  if (sum(valid) < 100) {
    return(c(NA, NA))
  }
  
  rr_star <- (x1_star[valid] / n1) / (x0_star[valid] / n0)
  
  quantile(rr_star, c(alpha / 2, 1 - alpha / 2), names = FALSE)
}

coverage_sim <- function(p0, rr, n, K = 10000, B = 5000, alpha = 0.05) {
  p1 <- rr * p0
  true_rr <- rr
  
  dm_cover <- 0L
  bs_cover <- 0L
  valid_dm <- 0L
  valid_bs <- 0L
  
  for (k in seq_len(K)) {
    x1 <- rbinom(1, n, p1)
    x0 <- rbinom(1, n, p0)
    
    if (x1 == 0 || x0 == 0) {
      next
    }
    
    dm_ci <- bls_rr_ci(x1, n, x0, n, alpha)
    valid_dm <- valid_dm + 1L
    dm_cover <- dm_cover + (dm_ci[1] <= true_rr & true_rr <= dm_ci[2])
    
    bs_ci <- bls_bootstrap_ci(x1, n, x0, n, B, alpha)
    if (!anyNA(bs_ci)) {
      valid_bs <- valid_bs + 1L
      bs_cover <- bs_cover + (bs_ci[1] <= true_rr & true_rr <= bs_ci[2])
    }
  }
  
  tibble(
    p0 = p0,
    rr = rr,
    n = n,
    ex0 = n * p0,
    dm_coverage = dm_cover / valid_dm,
    bs_coverage = bs_cover / valid_bs,
    valid_dm = valid_dm,
    valid_bs = valid_bs
  )
}
```

```{r}
#| label: run-sim
#| include: false
#| cache: true

set.seed(2026)

grid <- crossing(
  p0 = c(0.02, 0.05, 0.10, 0.20),
  n = c(100, 500, 2000),
  rr = c(1.20, 1.50)
) |>
  filter(rr == 1.20 | (p0 == 0.05 & n == 100))

results <- grid |>
  pmap_dfr(
    \(p0, n, rr) coverage_sim(p0, rr, n)
  ) |>
  arrange(ex0)
```

```{r}
#| label: tbl-coverage
#| tbl-cap: "Coverage of 95% CIs by condition. Both methods evaluated against the true $RR$ over $K = 10,000$ replications."
#| echo: false
#| message: false

results |>
  select(p0, n, rr, ex0, dm_coverage, bs_coverage) |>
  gt::gt() |>
  gt::fmt_number(columns = c(p0, rr), decimals = 2) |>
  gt::fmt_integer(columns = c(n, ex0)) |>
  gt::fmt_percent(columns = c(dm_coverage, bs_coverage), decimals = 1) |>
  gt::cols_label(
    p0 = gt::md("$p_0$"),
    n  = gt::md("$n$"),
    rr = gt::md("$RR$"),
    ex0 = gt::md("$\\mathbb{E}[x_0]$"),
    dm_coverage = "Delta method",
    bs_coverage = "Bootstrap"
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

```{r}
#| label: fig-coverage
#| fig-cap: "Coverage of 95% CIs as a function of $E[x_0]$. Dashed line: nominal 95%. Both methods overcover when $E[x_0]$ is small; the delta method more so."
#| echo: false
#| message: false

results |>
  filter(rr == 1.20) |>
  select(ex0, `Delta method` = dm_coverage, Bootstrap = bs_coverage) |>
  pivot_longer(-ex0, names_to = "method", values_to = "coverage") |>
  mutate(method = fct_inorder(method)) |>
  ggplot(aes(x = ex0, y = coverage, colour = method, shape = method)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.5) +
  geom_line() +
  scale_x_log10(breaks = c(2, 5, 10, 20, 50, 100, 200, 400)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    limits = c(0.93, 1.0)
  ) +
  labs(x = expression(E(x[0])), y = "Coverage", colour = NULL, shape = NULL) +
  theme_classic(base_family = "ubuntu", base_size = 13) +
  theme(legend.position = "top")
```

@fig-coverage and @tbl-coverage tell a clear story.

Large $\mathbb{E}[x_0]$: the methods agree. At $\mathbb{E}[x_0] \geq 50$, both methods land within about one percentage point of 95%. The delta method CI works as advertised.

Small $\mathbb{E}[x_0]$: the delta method overcovers. Below $\mathbb{E}[x_0] \approx 20$, coverage climbs above 95% for both methods, with the delta method rising faster. At $\mathbb{E}[x_0] = 5$, delta method coverage is around 97.5%. At $\mathbb{E}[x_0] = 2$, it exceeds 99%.

The mechanism is the $\frac{1}{x_0}$ term in $SE = \sqrt{\frac{1}{x_1} - \frac{1}{n_1} + \frac{1}{x_0} - \frac{1}{n_0}}$. When $x_0$ is 1, 2, or 3, this term dominates and inflates the SE, producing intervals far wider than the stated level implies. The CI is still valid (it covers at least 95% of the time) but conservative: the stated 95% does not reflect the actual coverage.

RR-independence confirmed. At $\mathbb{E}[x_0] = 5$, the $RR = 1.50$ confirmation condition produces coverage nearly identical to $RR = 1.20$, as predicted.

$\mathbb{E}[x_0]$ sufficiency. The duplicate pairs ($\mathbb{E}[x_0] = 10$ and $100$, each from two different $(p_0, n)$ combinations) show coverage differences under one percentage point. $\mathbb{E}[x_0]$ is the primary driver.

## Minimum sample size

The delta method CI is reliable when $\mathbb{E}[x_0]$ is large enough. The practical question: how many respondents per group does a study need to detect a given lift?

### From CI to power

A lift is detected when the CI lower bound exceeds 1. On the log scale:

$$\log(\widehat{RR}) - z_{1-\alpha/2} \cdot SE > 0$$

Under the true $RR$, $\log(\widehat{RR}) \sim N(\log(RR), \; SE^2)$ approximately. Power is the probability of detection:

$$\text{Power} = P\!\left(\log(\widehat{RR}) > z_{1-\alpha/2} \cdot SE\right) = \Phi\!\left(\frac{\log(RR)}{SE} - z_{1-\alpha/2}\right)$$

Setting power to $1 - \beta$ and solving:

$$\frac{\log(RR)}{SE} = z_{1-\alpha/2} + z_{1-\beta}$$


### Derivation of $n_{\min}$

Set $n_1 = n_0 = n$. From $\log(RR) = (z_{1-\alpha/2} + z_{1-\beta}) \cdot SE$ and $SE^2 = \frac{1-p_1}{n\,p_1} + \frac{1-p_0}{n\,p_0}$, solve for $n$:

$$n_{\min} = \left\lceil \frac{(z_{1-\alpha/2} + z_{1-\beta})^2}{[\log(RR)]^2}\left[\frac{1-p_1}{p_1} + \frac{1-p_0}{p_0}\right] \right\rceil$$

Since $p_1 = RR \cdot p_0$, the formula takes four inputs: $p_0$, $RR$, $\alpha$, and power ($1-\beta$).


### Worked example

Planning inputs from historical study norms: $p_0 = 0.10$, $RR = 1.20$ (20% relative lift). Target: 80% power at $\alpha = 0.05$. Then $p_1 = 0.12$, $z_{0.975} = 1.96$, $z_{0.80} = 0.842$, and:

$$n_{\min} = \left\lceil \frac{(1.96 + 0.842)^2}{[\log(1.20)]^2}\left[\frac{1-0.12}{0.12} + \frac{1-0.10}{0.10}\right] \right\rceil = \left\lceil \frac{7.849}{0.03324} \times 16.333 \right\rceil = \left\lceil 236.1 \times 16.333 \right\rceil = 3{,}857$$

At 80% power, the study needs approximately 3,900 respondents per group to detect a 20% relative lift.

$p_0$ and $RR$ are planning assumptions, typically from historical norms; if the actual values deviate, the realized power will differ.


### $n_{\min}$ reference table

```{r}
#| label: fun-nmin
#| include: false

bls_nmin <- function(p0, rr, alpha = 0.05, power = 0.80) {
  p1 <- rr * p0
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(power)
  ceiling(
    (z_alpha + z_beta)^2 / log(rr)^2
    * ((1 - p1) / p1 + (1 - p0) / p0)
  )
}
```

```{r}
#| label: tbl-nmin
#| tbl-cap: "Minimum respondents per group ($n_{\\min}$) to detect lift with 80% power at $\\alpha = 0.05$."
#| echo: false
#| message: false

nmin_grid <- crossing(
  p0 = c(0.05, 0.10, 0.20),
  rr = c(1.10, 1.20, 1.50)
) |>
  mutate(n_min = map2_int(p0, rr, \(p0, rr) bls_nmin(p0, rr)))

nmin_grid |>
  gt::gt() |>
  gt::fmt_number(columns = p0, decimals = 2) |>
  gt::fmt_number(columns = rr, decimals = 2) |>
  gt::fmt_integer(columns = n_min) |>
  gt::cols_label(
    p0 = gt::md("$p_0$"),
    rr = gt::md("$RR$"),
    n_min = gt::md("$n_{\\min}$")
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```


## R code

``` r
bls_rr_ci <- function(x1, n1, x0, n0, alpha = 0.05) {
  p1 <- x1 / n1
  p0 <- x0 / n0
  rr <- p1 / p0

  se_log <- sqrt(1/x1 - 1/n1 + 1/x0 - 1/n0)
  z <- qnorm(1 - alpha / 2)
  exp(log(rr) + c(-1, 1) * z * se_log)
}

bls_bootstrap_ci <- function(x1, n1, x0, n0, B = 10000, alpha = 0.05) {
  p1_hat <- x1 / n1
  p0_hat <- x0 / n0
  x1_star <- rbinom(B, n1, p1_hat)
  x0_star <- rbinom(B, n0, p0_hat)

  # Discard replicates where x0* = 0 (RR undefined)
  valid <- x0_star > 0 & x1_star > 0
  rr_star <- (x1_star[valid] / n1) / (x0_star[valid] / n0)
  quantile(rr_star, c(alpha / 2, 1 - alpha / 2))
}

bls_nmin <- function(p0, rr, alpha = 0.05, power = 0.80) {
  p1 <- rr * p0
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(power)
  ceiling(
    (z_alpha + z_beta)^2 / log(rr)^2
    * ((1 - p1) / p1 + (1 - p0) / p0)
  )
}

bls_rr_ci(105, 500, 70, 500)
# [1] 1.138381 1.976491

bls_bootstrap_ci(105, 500, 70, 500)
#     2.5%    97.5%
# 1.146667 1.984854

bls_nmin(p0 = 0.10, rr = 1.20)
# [1] 3857
```
