---
title: "From Risk Ratios to Odds Ratios: Choosing an Aggregation Scale for Brand Lift"
draft: true
draft-mode: unlinked
date: "2026-08-12"
categories: ["brand-lift", "statistics"]
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

The previous posts built confidence intervals for the risk ratio of a single Brand Lift Study (BLS). This post addresses a different question: how to compare and aggregate effects across studies with different baseline response rates. The scale used to report a single campaign's result to an advertiser and the scale used to pool results across studies serve different purposes and need not be the same.


## The ceiling effect: RR and absolute lift depend on the baseline

$RR = \frac{p_1}{p_0}$ is bounded above by $\frac{1}{p_0}$, since $p_1 \leq 1$. A study with $p_0 = 0.50$ can produce at most $RR = 2$. A study with $p_0 = 0.05$ can produce at most $RR = 20$. The maximum achievable $RR$ is a property of the baseline, not of the ad's effectiveness.

The absolute lift $p_1 - p_0$ has the same problem in the other direction: its maximum is $1 - p_0$. A study with $p_0 = 0.80$ can produce at most $p_1 - p_0 = 0.20$, regardless of how effective the ad is.

Both $RR$ and absolute lift carry information about the baseline baked into their numerical value. Comparing $RR = 1.30$ from one study to $RR = 1.15$ from another and concluding the first ad was more effective requires the assumption that the two baselines impose similar constraints, which is rarely the case across BLS studies.


## A baseline-free effect requires fixing a scale

Is there an effect measure that does not move with $p_0$? There is, but it requires choosing a scale on which the effect is defined to be constant. That choice is a model assumption.

The logistic regression model assumes ad exposure shifts the log-odds of a positive response by a constant $\beta$:

$$\text{logit}(p_1) = \text{logit}(p_0) + \beta$$

Under this model, $OR = \exp(\beta)$ is the same regardless of $p_0$. Given $p_0$ and $OR$, the exposed rate is:

$$p_1 = \frac{p_0 \cdot OR}{1 - p_0 + p_0 \cdot OR}$$

Plugging in $OR = 1.30$:

| $p_0$ | $p_1$  | $RR$  | $p_1 - p_0$  | $OR$ |
|--------|--------|-------|-------|------|
| 0.05   | 0.064  | 1.278 | 0.014 | 1.30 |
| 0.10   | 0.126  | 1.262 | 0.026 | 1.30 |
| 0.30   | 0.358  | 1.193 | 0.058 | 1.30 |
| 0.50   | 0.565  | 1.130 | 0.065 | 1.30 |

$OR$ is constant by construction. $RR$ decreases from 1.278 to 1.130 as $p_0$ increases: the ceiling compresses it. Absolute lift increases from 0.014 to 0.065. The same logit-scale effect produces different $RR$ and $p_1 - p_0$ values at different baselines.

This is a model assumption, not a fact. Whether ad exposure truly operates on the logit scale is untestable from a single 2×2 table. The assumption is adopted because it makes cross-study comparison coherent, not because it is known to be correct.


## The odds ratio and the variance of log(OR)

The odds ratio is defined as:

$$OR = \frac{p_1 / (1 - p_1)}{p_0 / (1 - p_0)}$$

Label the four cells of the 2×2 table as $a = x_1$, $b = n_1 - x_1$, $c = x_0$, $d = n_0 - x_0$. Then:

$$\widehat{OR} = \frac{a \cdot d}{b \cdot c}$$

$$\log(\widehat{OR}) = \text{logit}(\hat{p}_1) - \text{logit}(\hat{p}_0)$$

Apply the delta method to $g(p) = \text{logit}(p) = \log(p) - \log(1-p)$. The derivative is $g'(p) = \frac{1}{p(1-p)}$, so:

$$\text{Var}(\text{logit}(\hat{p})) \approx \frac{1}{[p(1-p)]^2} \cdot \frac{p(1-p)}{n} = \frac{1}{n \, p(1-p)}$$

Rewriting with cell counts $x = np$ and $n - x = n(1-p)$:

$$\text{Var}(\text{logit}(\hat{p})) = \frac{1}{x} + \frac{1}{n - x}$$

Since $\hat{p}_1$ and $\hat{p}_0$ are independent, the variance of the difference is the sum:

$$\text{Var}(\log(\widehat{OR})) = \frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}$$

The [first post](/posts/bls-delta-method-rr/) derived $\text{Var}(\log(\widehat{RR})) = \frac{1}{x_1} - \frac{1}{n_1} + \frac{1}{x_0} - \frac{1}{n_0}$, which can be rewritten as $\frac{1}{a} - \frac{1}{a+b} + \frac{1}{c} - \frac{1}{c+d}$. The variance of $\log(\widehat{OR})$ has a simpler form: four reciprocals with no subtraction.


## Constructing the confidence interval for OR

The CI for $\log(OR)$ follows the same structure as the log-transform CI for $RR$ from the first post: symmetric on the log scale, exponentiated back.

$$\log(\widehat{OR}) \pm z_{1-\alpha/2} \times \sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}$$

Exponentiating:

$$\left( \widehat{OR} \cdot \exp(-z_{1-\alpha/2} \cdot SE), \quad \widehat{OR} \cdot \exp(+z_{1-\alpha/2} \cdot SE) \right)$$

where $SE = \sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}$.

The interval is always positive, asymmetric around $\widehat{OR}$, and equivariant under inversion of the table.


## Two properties that distinguish log(OR) from log(RR)

Both $\log(\widehat{RR})$ and $\log(\widehat{OR})$ have closed-form variance formulas derived from the delta method. Both support log-transform CIs, SQL implementation, and inverse-variance weighting. The closed-form variance is not a differentiator. Two properties are.

**Baseline independence under logit additivity.** If ad exposure shifts log-odds by a constant $\beta$, then $OR = \exp(\beta)$ does not depend on $p_0$. $RR$ and $p_1 - p_0$ do. This is the reason to use $\log(OR)$ for cross-study aggregation: pooling $\log(OR)$ values from studies with different baselines estimates a common effect on the logit scale. Pooling $\log(RR)$ values conflates the ad effect with the baseline.

**Transpose symmetry.** Swapping rows (exposed $\leftrightarrow$ control) or columns (positive $\leftrightarrow$ negative) in the 2×2 table maps $\log(OR) \to -\log(OR)$: the sign flips, the magnitude is unchanged. $\text{Var}(\log(\widehat{OR})) = \frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}$ is symmetric in all four cells, reflecting this property. $\text{Var}(\log(\widehat{RR})) = \frac{1}{a} - \frac{1}{a+b} + \frac{1}{c} - \frac{1}{c+d}$ is not symmetric: swapping positive and negative produces a different variance, because $RR$ does not have transpose symmetry.


## When the odds ratio approximates the risk ratio

The two measures are related by an identity:

$$OR = RR \times \frac{1 - p_0}{1 - p_1}$$

When both $p_0$ and $p_1$ are small, $\frac{1 - p_0}{1 - p_1} \approx 1$ and $OR \approx RR$.

BLS baselines are not small. Typical values range from 10% to 40%. At these levels, $OR$ and $RR$ diverge:

| $p_0$ | $p_1$ | $RR$  | $OR$  |
|--------|--------|-------|-------|
| 0.02   | 0.024  | 1.200 | 1.205 |
| 0.10   | 0.120  | 1.200 | 1.227 |
| 0.30   | 0.360  | 1.200 | 1.313 |

At $p_0 = 0.30$ with $RR = 1.20$, the $OR$ is $1.31$, a 55% overstatement of the lift if misread as a risk ratio. In BLS contexts, $OR$ must not be interpreted as lift. The reporting scale for advertisers remains $RR - 1$ (relative lift) or $p_1 - p_0$ (absolute lift). $\log(OR)$ serves a different purpose: aggregation.


## Worked example: three studies at different baselines

Fix $\beta = \log(1.30) = 0.2624$ and construct three synthetic studies with $n_1 = n_0 = 500$. Cell counts are $n \times p$ rounded to the nearest integer.

```{r}
#| label: tbl-studies
#| tbl-cap: "Three synthetic studies with constant log-odds effect $\\beta = \\log(1.30)$. OR is approximately constant; RR decreases with higher baseline."
#| echo: false
#| message: false

studies <- tribble(
  ~study, ~p0,  ~a,  ~b,   ~c,  ~d,
  "A",    0.10,  63, 437,  50, 450,
  "B",    0.30, 179, 321, 150, 350,
  "C",    0.50, 283, 217, 250, 250
) |>
  mutate(
    p1 = a / (a + b),
    rr = p1 / p0,
    or = (a * d) / (b * c)
  )

studies |>
  select(study, p0, p1, a, b, c, d, rr, or) |>
  gt::gt() |>
  gt::fmt_number(columns = c(p0, p1), decimals = 3) |>
  gt::fmt_number(columns = c(rr, or), decimals = 3) |>
  gt::fmt_integer(columns = c(a, b, c, d)) |>
  gt::cols_label(
    study = "Study",
    p0 = gt::md("$p_0$"),
    p1 = gt::md("$p_1$"),
    a = gt::md("$a$"),
    b = gt::md("$b$"),
    c = gt::md("$c$"),
    d = gt::md("$d$"),
    rr = gt::md("$RR$"),
    or = gt::md("$OR$")
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

The log-odds effect is the same across all three studies by construction. $RR$ ranges from 1.132 (study C, $p_0 = 0.50$) to 1.260 (study A, $p_0 = 0.10$). Ranking by $RR$ would make study A appear most effective. Ranking by $OR$ gives approximately the same value for all three.

The per-study CIs for $OR$:

```{r}
#| label: tbl-study-ci
#| tbl-cap: "Per-study 95% CIs for OR via the log-transform delta method."
#| echo: false
#| message: false

study_ci <- studies |>
  mutate(
    log_or = log(or),
    se2 = 1/a + 1/b + 1/c + 1/d,
    se = sqrt(se2),
    ci_lower = exp(log_or - 1.96 * se),
    ci_upper = exp(log_or + 1.96 * se)
  )

study_ci |>
  select(study, log_or, se2, se, ci_lower, ci_upper) |>
  gt::gt() |>
  gt::fmt_number(columns = c(log_or, se2, se), decimals = 4) |>
  gt::fmt_number(columns = c(ci_lower, ci_upper), decimals = 3) |>
  gt::cols_label(
    study = "Study",
    log_or = gt::md("$\\log(\\widehat{OR})$"),
    se2 = gt::md("$SE^2$"),
    se = gt::md("$SE$"),
    ci_lower = "CI lower",
    ci_upper = "CI upper"
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

Study A has the widest CI ($SE = 0.201$), driven by the small control positive count $c = 50$: the $\frac{1}{c}$ term dominates. Study C has the narrowest ($SE = 0.127$), because all four cells are large. Only study C's CI excludes $1$ at the 95% level.


## Aggregation: why not merge the tables?

The three 2×2 tables could be summed cell-by-cell into a single pooled table. If the $OR$ is the same within each study, the pooled $OR$ should recover that common value. It does not. A numerical example with two studies, each with $OR = 4.0$ and $n_1 = n_0 = 1{,}000$:

```{r}
#| label: tbl-noncollapse
#| tbl-cap: "Non-collapsibility of the OR. Within-study OR is 4.0 for both studies. The marginal OR is 3.22."
#| echo: false
#| message: false

nc <- tribble(
  ~study, ~p0,  ~p1,
  "A",    0.10,  4/13,
  "B",    0.40,  8/11,
) |>
  mutate(
    or = (p1 / (1 - p1)) / (p0 / (1 - p0)),
  )

nc_merged <- nc |>
  summarise(
    study = "Merged",
    p0 = mean(p0),
    p1 = mean(p1),
  ) |>
  mutate(
    or = (p1 / (1 - p1)) / (p0 / (1 - p0)),
  )

bind_rows(nc, nc_merged) |>
  gt::gt() |>
  gt::fmt_number(columns = c(p0, p1, or), decimals = 3) |>
  gt::cols_label(
    study = "Study",
    p0 = gt::md("$p_0$"),
    p1 = gt::md("$p_1$"),
    or = gt::md("$OR$")
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

Both studies have $OR = 4.0$, but the merged table gives $OR = 3.22$. Summing the tables distorts the $OR$, even though the within-study effect is the same. ($RR$ does not have this problem: merging preserves a common $RR$.) The fix is straightforward: compute $\log(\widehat{OR})$ within each study, then pool the estimates.


## Pooling study-level log(OR)

The inverse-variance weighted mean of $\log(\widehat{OR})$ across $K$ studies is:

$$\log(\widehat{OR})_{\text{pooled}} = \frac{\sum_{k=1}^{K} w_k \log(\widehat{OR}_k)}{\sum_{k=1}^{K} w_k}$$

where $w_k = \frac{1}{\text{Var}(\log(\widehat{OR}_k))} = \frac{1}{\frac{1}{a_k} + \frac{1}{b_k} + \frac{1}{c_k} + \frac{1}{d_k}}$.

The variance of the pooled estimate is $\text{Var}(\log(\widehat{OR})_{\text{pooled}}) = \frac{1}{\sum_{k} w_k}$, and the 95% CI is:

$$\log(\widehat{OR})_{\text{pooled}} \pm 1.96 \times \frac{1}{\sqrt{\sum_k w_k}}$$


### Worked example: pooling the three studies

```{r}
#| label: tbl-pooling
#| tbl-cap: "Inverse-variance weighted pooling of three studies."
#| echo: false
#| message: false

pooling <- study_ci |>
  mutate(
    w = 1 / se2,
    w_log_or = w * log_or
  )

w_total <- sum(pooling$w)
pooled_log_or <- sum(pooling$w_log_or) / w_total
pooled_se <- 1 / sqrt(w_total)
pooled_ci_lower <- exp(pooled_log_or - 1.96 * pooled_se)
pooled_ci_upper <- exp(pooled_log_or + 1.96 * pooled_se)

pooling |>
  select(study, log_or, se2, w, w_log_or) |>
  gt::gt() |>
  gt::fmt_number(columns = c(log_or, se2, w, w_log_or), decimals = 4) |>
  gt::cols_label(
    study = "Study",
    log_or = gt::md("$\\log(\\widehat{OR})$"),
    se2 = gt::md("$SE^2$"),
    w = gt::md("$w$"),
    w_log_or = gt::md("$w \\times \\log(\\widehat{OR})$")
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

$$\sum w_k = 141.58, \quad \log(\widehat{OR})_{\text{pooled}} = \frac{37.3433}{141.58} = 0.2638$$

$$SE_{\text{pooled}} = \frac{1}{\sqrt{141.58}} = 0.0840$$

$$\text{95\% CI for } \log(OR): \quad 0.2638 \pm 1.96 \times 0.0840 = (0.0990, \; 0.4285)$$

$$\text{95\% CI for } OR: \quad (e^{0.0990}, \; e^{0.4285}) = (1.104, \; 1.535)$$

None of the three individual studies produced a CI that cleanly excluded $1$ at the 95% level (only study C barely did). The pooled CI excludes $1$, with a lower bound above $1.1$. Aggregation recovered the signal that the individual studies could not establish alone.

Study C receives the largest weight ($w_C$) because its four cells are all large, producing the smallest variance. Study A receives the smallest weight: its control positive count $c = 50$ inflates the $\frac{1}{c}$ term. The weights automatically downweight less precise estimates, without any manual judgment about which studies to include.

```{r}
#| label: fig-pooled-ci
#| fig-cap: "Per-study and pooled 95% CIs for OR."
#| echo: false
#| message: false

ci_data <- study_ci |>
  select(label = study, or, ci_lower, ci_upper) |>
  bind_rows(
    tibble(
      label = "Pooled",
      or = exp(pooled_log_or),
      ci_lower = pooled_ci_lower,
      ci_upper = pooled_ci_upper
    )
  ) |>
  mutate(label = fct_inorder(label) |> fct_rev())

ggplot(ci_data, aes(x = or, y = label, xmin = ci_lower, xmax = ci_upper, colour = label)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_pointrange() +
  scale_x_continuous(breaks = scales::breaks_width(0.25)) +
  labs(x = "Odds Ratio", y = NULL) +
  theme_classic(base_family = "ubuntu", base_size = 13) +
  theme(legend.position = "none")
```
