---
title: >-
  Bayesian Structural Time Series (BSTS): From Point Estimates to Posterior Predictive Intervals
draft: true
draft-mode: unlinked
date: "2026-11-25"
categories: ["causal-inference", "statistics"]
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

date_treat <- lubridate::ymd("2026-06-01")
att_true <- 0.01049 # True ATT derived from the data-generating process (+1.049 pp)

panel <- read_csv(
  "../../data/causal-synthetic-data-design/panel_parallel.csv",
  show_col_types = FALSE
)
```

This article models the temporal structure of the iOS completion rate with a state-space model and estimates the causal effect by predicting how the rate would have evolved without the intervention.
The method is Bayesian structural time series (BSTS), implemented with {CausalImpact} from [Brodersen et al. (2015)](https://doi.org/10.1214/14-AOAS788).

The data are daily ad completion rates for 11 devices over 243 days, generated in the [article on synthetic data design](/posts/causal-synthetic-data-design/).
An ad-pause feature was released only on iOS on 2026-06-01.
The true average treatment effect on the treated (ATT) is +1.049 percentage points (pp).
To evaluate the method, I use the full 182-day pre-intervention period and 61-day post-intervention period.

This article covers the following:

1. Define the BSTS state-space model and the selection of control series through a spike-and-slab prior.
2. Estimate pointwise effects and the post-intervention average with {CausalImpact}.
3. Compare the results and assumptions of difference-in-differences (DiD), the [synthetic control method (SCM)](/posts/causal-scm-placebo-inference/), and CausalImpact.

## State-space model

BSTS combines the treated unit's own temporal structure with a regression on control series.
A local level and a day-of-week component represent the temporal structure of iOS, while the regression component represents variation shared with the controls.

Let $y_t$ denote the daily iOS completion rate and $\mathbf{x}_t$ the vector of completion rates for the 10 controls.
The index $t = 1, \dots, 243$ corresponds to dates from 2025-12-01 through 2026-07-31, and the release date $T$ = 2026-06-01 corresponds to $t = 183$.
The observation equation is

$$
  y_t = \mu_t + \gamma_t + \boldsymbol{\beta}^\top \mathbf{x}_t
        + \varepsilon_t, \qquad \varepsilon_t \sim N(0, \sigma_\varepsilon^2)
$$

  | Component                                                   | State transition                                                                                    | Variation represented                                                  |
  | ----------------------------------------------------------- | --------------------------------------------------------------------------------------------------- | ---------------------------------------------------------------------- |
  | Local level $\mu_t$                                         | $\mu_{t+1} = \mu_t + \eta_t, \quad \eta_t \sim N(0, \sigma_\eta^2)$                                 | Local variation in the iOS level                                       |
  | Day-of-week component $\gamma_t$                            | $\gamma_{t+1} = -\sum_{s=0}^{5} \gamma_{t-s} + \omega_t, \quad \omega_t \sim N(0, \sigma_\omega^2)$ | The iOS weekly cycle (expected to sum to 0 over seven days)            |
  | Regression component $\boldsymbol{\beta}^\top \mathbf{x}_t$ | None ($\boldsymbol{\beta}$ is time-invariant)                                                       | Variation shared with the controls, including unobserved common causes |

The control series must not be affected by the intervention.
Because the intervention changes the UI within the iOS app, spillovers to other devices are unlikely.

I place a spike-and-slab prior on the regression coefficients $\boldsymbol{\beta}$.
This mixture of a point mass at 0 (the spike) and a diffuse continuous distribution (the slab) represents uncertainty about whether each control series enters the regression.
SCM also determines control weights from the data, but the roles and constraints differ.

  | Aspect                       | Convex combination in SCM                                                     | Regression component in BSTS                                  |
  | ---------------------------- | ----------------------------------------------------------------------------- | ------------------------------------------------------------- |
  | Source of the counterfactual | Donor series only                                                             | The treated unit's state components + control series          |
  | Coefficient constraints      | $w_j \geq 0$, $\sum_j w_j = 1$                                                | None (sign and sum are unrestricted)                          |
  | Level differences            | No intercept; reproduces the level within the convex hull                     | Absorbed by the local level $\mu_t$                           |
  | Treatment of control series  | Point-estimated weights from constrained optimization (some weights may be 0) | Probabilistic selection through a spike-and-slab prior        |
  | Selection uncertainty        | Not represented (one point estimate of the weights)                           | Represented by the posterior distribution of the coefficients |

BSTS has no convex-combination constraint, and $\mu_t$ absorbs level differences, so the treated unit need not lie within the donors' convex hull.
The tradeoff is that extrapolation is allowed: predictions outside the range observed in the pre-intervention period depend strongly on the model structure and prior.

## Posterior prediction of the counterfactual

The posterior distribution of the model is estimated from $(y_t, \mathbf{x}_t)$ over the 182-day pre-intervention period.
For the 61-day post-intervention period, the predictive distribution of $y_t^{(0)}$ is computed using only the observed control series $\mathbf{x}_t$.
This distribution is the counterfactual completion rate without the intervention.
{CausalImpact} fits the model over the full period while treating post-intervention $y_t$ as missing, then extracts its posterior predictive distribution.
This prediction assumes that the regression coefficients $\boldsymbol{\beta}$ and the structure of the state components remain stable after the intervention.
Like parallel trends in DiD or the assumption in SCM that the pre-intervention factor structure continues afterward, this assumption cannot be verified from pre-intervention fit alone.

Define the pointwise effect as $\tau_t = y_t - y_t^{(0)}$ and the post-intervention average as

$$
  \bar{\tau} = \frac{1}{61}\sum_{t=183}^{243}\tau_t
$$

For each posterior predictive draw of the counterfactual $y_t^{(0)}$, I calculate its difference from the observation and the average of those differences over the period.
This yields the posterior distributions of $\tau_t$ and $\bar{\tau}$.
I report the interval from the 2.5th to the 97.5th percentile as the 95% posterior interval.

In principle, the interval for the pointwise effect widens with distance from the release date because $\sigma_\eta^2$ is added to the predictive variance of the local level each day.
If $\sigma_\eta$ is sufficiently small relative to the observation noise $\sigma_\varepsilon$, however, little widening may be visible over 61 days.

{CausalImpact} reports pointwise effects, the post-intervention average, and a cumulative effect (`cumulative`).
I report only $\tau_t$ and $\bar{\tau}$.
A sum of completion rates with denominators that vary by day has no meaningful interpretation, so I do not use the cumulative effect.

## Estimation with {CausalImpact}

The default {CausalImpact} model contains a local level and a regression component but no seasonal component (`nseasons = 1`).
Because these data have a day-of-week cycle, I set `nseasons = 7`.

The model standardizes each series using its pre-intervention mean and standard deviation, fits the model, and transforms the results back to the original scale.
I retain the default priors.
For the standard deviation $\sigma_\eta$ of the local-level disturbance $\eta_t$, the defaults specify a guess of 0.01 on the standardized scale (`prior.level.sd`) and 32 prior observations, implying a slowly varying level.
The spike-and-slab prior has an expected model size of 3, giving each of the 10 control series a prior inclusion probability of 0.3.
I do not examine the other settings ([{CausalImpact} source](https://github.com/google/CausalImpact/blob/master/R/impact_model.R)).

The first column of the input data is the date, the second is iOS, and the remaining columns are the 10 controls.
The pre- and post-intervention periods are specified by date.
{CausalImpact} internally fixes the {bsts} random seed at 1.
I draw 5,000 Markov chain Monte Carlo (MCMC) samples.

```{r}
#| label: did-61
#| include: false

# Compute the same 2x2 DiD estimator (iOS vs Android) over 61 post-intervention days
did_61 <- panel |>
  filter(device %in% c("ios", "android")) |>
  mutate(period = if_else(post, "post", "pre")) |>
  summarise(
    complete = sum(complete),
    impression = sum(impression),
    .by = c(device, period)
  ) |>
  mutate(
    rate = complete / impression,
    var_rate = rate * (1 - rate) / impression
  ) |>
  pivot_wider(
    names_from = period,
    values_from = c(rate, var_rate),
    id_cols = device
  ) |>
  mutate(
    diff = rate_post - rate_pre,
    var_diff = var_rate_post + var_rate_pre,
    .keep = "unused"
  ) |>
  pivot_wider(names_from = device, values_from = c(diff, var_diff)) |>
  mutate(
    estimate = diff_ios - diff_android,
    se = sqrt(var_diff_ios + var_diff_android),
    ci_lower = estimate - qnorm(0.975) * se,
    ci_upper = estimate + qnorm(0.975) * se,
    .keep = "none"
  )
```

```{r}
#| label: impact-fit
#| include: false
#| cache: true

rates <- panel |>
  mutate(device = str_replace_all(device, "-", "_")) |>
  pivot_wider(
    id_cols = date,
    names_from = device,
    values_from = complete_rate
  ) |>
  arrange(date) |>
  relocate(date, ios) # First column: date; second column: y (iOS)

pre_period <- c(min(rates$date), date_treat - 1)
post_period <- c(date_treat, max(rates$date))

# Local level + day-of-week (nseasons = 7) + regression
impact <- CausalImpact::CausalImpact(
  rates,
  pre.period = pre_period,
  post.period = post_period,
  model.args = list(nseasons = 7, niter = 5000)
)

# Reference model without seasonality (used only to compare inclusion probabilities)
impact_noseason <- CausalImpact::CausalImpact(
  rates,
  pre.period = pre_period,
  post.period = post_period,
  model.args = list(niter = 5000)
)

avg <- impact$summary |>
  as_tibble(rownames = "period") |>
  pivot_longer(-period, names_to = "metric") |>
  filter(period == "Average")
```

```{r}
#| label: tbl-impact
#| tbl-cap: >-
#|   Post-intervention averages.
#|   Observed, counterfactual, and effect values are shown in pp.
#| echo: false
#| message: false

avg |>
  pivot_wider(names_from = metric) |>
  select(
    Actual,
    Pred,
    Pred.lower,
    Pred.upper,
    AbsEffect,
    AbsEffect.lower,
    AbsEffect.upper
  ) |>
  gt::gt() |>
  gt::fmt_number(decimals = 3, scale_by = 100) |>
  gt::cols_label(
    Actual = "Observed",
    Pred = "Posterior mean",
    Pred.lower = "Lower",
    Pred.upper = "Upper",
    AbsEffect = "Posterior mean",
    AbsEffect.lower = "Lower",
    AbsEffect.upper = "Upper"
  ) |>
  gt::tab_spanner(
    label = "Counterfactual",
    columns = c(Pred, Pred.lower, Pred.upper)
  ) |>
  gt::tab_spanner(
    label = gt::md("Effect $\\bar{\\tau}$"),
    columns = c(AbsEffect, AbsEffect.lower, AbsEffect.upper)
  ) |>
  gt::cols_align(align = "right") |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

@tbl-impact shows the post-intervention averages.
The difference between the posterior mean of $\bar{\tau}$ and the true value of +1.049 pp is `r scales::number(100 * (avg$AbsEffect - att_true), accuracy = 0.001, style_positive = "plus")` pp, and the 95% posterior interval contains the true value.

```{r}
#| label: fig-impact
#| fig-cap: >-
#|   Top: observed iOS completion rate (solid), posterior prediction of the counterfactual (dashed), and 95% posterior predictive interval (band).
#|   Bottom: pointwise effect $\tau_t$ and 95% posterior interval.
#|   The vertical dashed line marks the release date.
#| echo: false
#| message: false
#| warning: false

plot(impact, c("original", "pointwise")) +
  theme_classic(base_family = "ubuntu", base_size = 13) +
  theme(strip.background = element_blank())
```

In the top panel of @fig-impact, the observed series exceeds the counterfactual after release.
The interval width in the bottom panel is nearly constant over the 61 days because $\sigma_\eta$ for the local level is small relative to the observation noise, so the increase in variance over the prediction horizon is not conspicuous.
The pointwise intervals include daily binomial noise (with a standard deviation of about 0.4 pp) and are therefore wider than the interval for the average effect.
The figure is used to assess whether the effect persists over the period, not to judge individual days.

```{r}
#| label: tbl-inclusion
#| tbl-cap: >-
#|   Posterior inclusion probabilities for the regression coefficients (share of post-burn-in draws in which the coefficient is nonzero; first 10% discarded).
#|   The prior inclusion probability is 0.3 in both models.
#| echo: false
#| message: false

burn <- tibble(
  model = c("seasonal", "noseason"),
  burn = c(
    bsts::SuggestBurn(0.1, impact$model$bsts.model),
    bsts::SuggestBurn(0.1, impact_noseason$model$bsts.model)
  )
)

draws <- bind_rows(
  seasonal = as_tibble(impact$model$bsts.model$coefficients),
  noseason = as_tibble(impact_noseason$model$bsts.model$coefficients),
  .id = "model"
) |>
  mutate(draw = row_number(), .by = model) |>
  left_join(burn, by = "model") |>
  filter(draw > burn) |>
  select(-burn)

inclusion <- draws |>
  pivot_longer(-c(model, draw), names_to = "series", values_to = "beta") |>
  filter(series != "(Intercept)") |>
  summarise(inclusion = mean(beta != 0), .by = c(model, series)) |>
  mutate(series = str_replace_all(series, "_", "-"))

inclusion |>
  pivot_wider(names_from = model, values_from = inclusion) |>
  arrange(desc(noseason)) |>
  gt::gt() |>
  gt::fmt_percent(columns = -series, decimals = 1) |>
  gt::cols_label(
    series = "Control series",
    seasonal = "With day-of-week",
    noseason = "Without day-of-week (default)"
  ) |>
  gt::cols_align(align = "right", columns = -series) |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

@tbl-inclusion shows the posterior inclusion probabilities of the regression coefficients.
In the model with a day-of-week component, all series range from `r scales::percent(min(inclusion$seasonal), accuracy = 0.1)` to `r scales::percent(max(inclusion$seasonal), accuracy = 0.1)`, well below the prior inclusion probability of 0.3.

This agrees with the data-generating process.
Only the day-of-week cycle and trend are common across devices; daily binomial noise is independent.
Once the day-of-week component and local level capture the former, little shared variation remains for the control series to explain.
Without the day-of-week component, the inclusion probability for Android reaches `r scales::percent(inclusion$noseason[inclusion$series == "android"], accuracy = 0.1)`, and the regression component captures the day-of-week cycle instead.

SCM must use controls to form a convex combination, whereas BSTS selects controls only when they provide shared variation not explained by the state components.
Low inclusion probabilities in real data indicate that the counterfactual relies mainly on extrapolating the treated unit's own temporal structure.

## Comparison of three methods

I compare the three methods over 61 post-intervention days.
The [DiD article](/posts/causal-did-parallel-trends/) used a seven-day window to match the operational decision, but here I recompute the same estimator over 61 days.
The SCM value comes from the [article on that method](/posts/causal-scm-placebo-inference/).

```{r}
#| label: tbl-comparison
#| tbl-cap: >-
#|   Post-intervention average estimates (pp) from three methods on the same data (seed = 1).
#|   The true ATT is +1.049 pp.
#| echo: false
#| message: false

estimates <- bind_rows(
  "DiD (61-day window)" = did_61 |>
    select(estimate, lower = ci_lower, upper = ci_upper),
  "SCM" = tibble(estimate = 0.0108),
  "CausalImpact" = avg |>
    pivot_wider(names_from = metric, values_from = value) |>
    select(
      estimate = AbsEffect,
      lower = AbsEffect.lower,
      upper = AbsEffect.upper
    ),
  .id = "method"
)

notes <- tribble(
  ~method               , ~interval                                                                    , ~assumption                                                                                ,
  "DiD (61-day window)" , "Asymptotic normal 95% CI based on binomial variance"                        , "Parallel trends"                                                                          ,
  "SCM"                 , "None (in-space placebo rank, p = 1/11)"                                     , "The pre-intervention factor structure continues afterward"                                ,
  "CausalImpact"        , "95% posterior interval based on posterior prediction of the counterfactual" , "The state structure and relationship with controls remain stable across the intervention"
)

comparison <- estimates |> left_join(notes, by = "method")

comparison |>
  gt::gt() |>
  gt::fmt_number(
    columns = c(estimate, lower, upper),
    rows = method != "SCM",
    decimals = 3,
    scale_by = 100
  ) |>
  gt::fmt_number(
    columns = estimate,
    rows = method == "SCM",
    decimals = 2,
    scale_by = 100
  ) |>
  gt::sub_missing(columns = c(lower, upper), missing_text = "—") |>
  gt::cols_label(
    method = "Method",
    estimate = "Estimate",
    lower = "Lower",
    upper = "Upper",
    interval = "Inference",
    assumption = "Assumption supporting the counterfactual"
  ) |>
  gt::cols_align(align = "right", columns = c(estimate, lower, upper)) |>
  gt::opt_table_font(font = gt::google_font("Lora"))
```

All three point estimates are close to the true value because the data-generating process was designed to satisfy each method's assumptions.
In real data, parallel trends, persistence of the factor structure, and stability of the relationship with controls cannot be verified for the post-intervention period.
Pre-intervention trends and model fit are informative, but they do not prove these assumptions.

The CausalImpact counterfactual is composed almost entirely of the local level and day-of-week component.
Thus, this estimate is supported less by the relationship with controls than by the assumption that the temporal structure of iOS continues after the intervention.

The meaning of the inference also differs.
DiD uses an asymptotic normal CI based on binomial variance, SCM uses a rank with resolution $1/11$, and CausalImpact uses a posterior interval conditional on the model and prior.
All three nevertheless construct the counterfactual from assumptions that cannot be verified.

## R code

```r
library(tidyverse)

date_treat <- lubridate::ymd("2026-06-01")
att_true <- 0.01049

panel <- read_csv("../../data/causal-impact-bsts/panel_parallel.csv")

# Replace hyphens in device names because they are awkward in formulas
# First column: date; second: treated unit (iOS); remaining columns: 10 controls
rates <- panel |>
  mutate(device = str_replace_all(device, "-", "_")) |>
  pivot_wider(
    id_cols = date,
    names_from = device,
    values_from = complete_rate
  ) |>
  arrange(date) |>
  relocate(date, ios) # First column: date; second column: y (iOS)

pre_period <- c(min(rates$date), date_treat - 1)
post_period <- c(date_treat, max(rates$date))

# --- Local level + day-of-week + regression (CausalImpact internally fixes seed at 1) ---

impact <- CausalImpact::CausalImpact(
  rates,
  pre.period = pre_period,
  post.period = post_period,
  model.args = list(nseasons = 7, niter = 5000)
)

# Post-intervention average effect and 95% posterior interval (excluding cumulative row)
impact$summary["Average", c("AbsEffect", "AbsEffect.lower", "AbsEffect.upper")]

# Top: observed and counterfactual; bottom: pointwise effect (excluding cumulative panel)
plot(impact, c("original", "pointwise"))

# Posterior inclusion probabilities of regression coefficients (discarding first 10%)
burn <- bsts::SuggestBurn(0.1, impact$model$bsts.model)
colMeans(
  impact$model$bsts.model$coefficients[-seq_len(burn), , drop = FALSE] != 0
)

# --- Reference: default model without day-of-week component ---

impact_noseason <- CausalImpact::CausalImpact(
  rates,
  pre.period = pre_period,
  post.period = post_period,
  model.args = list(niter = 5000)
)

burn <- bsts::SuggestBurn(0.1, impact_noseason$model$bsts.model)
colMeans(
  impact_noseason$model$bsts.model$coefficients[
    -seq_len(burn),
    ,
    drop = FALSE
  ] !=
    0
)

# --- Comparison: DiD (iOS vs Android) over a 61-day post-intervention window ---

panel |>
  filter(device %in% c("ios", "android")) |>
  mutate(period = if_else(post, "post", "pre")) |>
  summarise(
    complete = sum(complete),
    impression = sum(impression),
    .by = c(device, period)
  ) |>
  mutate(
    rate = complete / impression,
    var_rate = rate * (1 - rate) / impression
  ) |>
  pivot_wider(
    names_from = period,
    values_from = c(rate, var_rate),
    id_cols = device
  ) |>
  mutate(
    diff = rate_post - rate_pre,
    var_diff = var_rate_post + var_rate_pre,
    .keep = "unused"
  ) |>
  pivot_wider(names_from = device, values_from = c(diff, var_diff)) |>
  mutate(
    estimate = diff_ios - diff_android,
    se = sqrt(var_diff_ios + var_diff_android),
    ci_lower = estimate - qnorm(0.975) * se,
    ci_upper = estimate + qnorm(0.975) * se,
    .keep = "none"
  )
```
